pirouette: the error BEAST2 makes in inferring a phylogeny

### 2019-02-19 article.tex
Updates from Overleaf

The development of new powerful inference tools, 
such as BEAST [\cite{drummond2007beast}] or MrBayes [\cite{huelsenbeck2001mrbayes}], 
provides us the possibility to build phylogenetic trees 
from genetic material extracted from extant organisms.
This has constituted a big step forward 
in doing experiments to elucidate on how species 
evolve.
Having access to this new kind of tools has given rise 
to a high interest in understanding what are the main drivers 
and modes for diversification processes.

There are plenty of diversification models: 
constant-birth [Yule, 19..],
birth-death [\cite{nee1994reconstructed}], 
time-dependent [Rabosky D., LASER package?], 
diversity dependent [\cite{etienne2011diversity}], 
protracted birth-death [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}],
multiple-birth death [Laudanno, Bilderbeek, Haegeman \& Etienne, in preparation],
a rate shifting birth-death model [the SLS model. Laudanno, Haegeman \& Etienne, in preparation]
punctuated equilibrium aka early burst [?Harmon],
dependent on a binary trait [\cite{maddison2007estimating}], 
multiple state trait [\cite{fitzjohn2012diversitree}],
concealed state [\cite{beaulieu2016detecting}] 
or even a combination of multiple concealed and observable states [\cite{herrera2018detecting}].

Such models usually rely on the assumption 
that a particular diversification process 
can be explained mainly by a specific mechanism.
Such a mechanism is built into a likelihood function, 
which provides the likelihood of the parameters, given the data.
By maximizing the likelihood, it is possible 
to estimate the most likely model parameters. 
It is a standard practice in assessing such models,
how well they estimate their own parameters
on simulated phylogenies in which the parameters are known.

Diversification models are getting more complex over time. There are many
reasons for this. One reason is the increase in computing power,
which allows this heightened complexity possible.
Another reason is to explore a process that is not
yet captured by the other diversification models. An example of this
is the multiple-birth model [Laudanno, Bilderbeek, Haegeman \& Etienne, in preparation], 
that is the first model to allow speciation to co-occur.

An open question for each new diversification model 
is whether the increased complexity is worth it. 
One of the purposes of modelling a complex systems
is to hypothesize the main driving factors of a process 
and trying to prove that they are indeed able 
to explain the main features of the phenomenon you want to explain.
Due to this, there are diminishing returns on adding complexity
to a model: the job of a modeller is to try to 
disentangle the intrinsic complexity of a system 
highlighting the main reasons of its behaviour. 
In such a light you don't want your model to be unnecessary complex.
Additionally, developing a likelihood model takes time, 
human resources, fundings and can strain computational resources.

When assuming that a new diversification model is worth the
(philosophical) effort, there are many ways to measure
if the new model indeed gives a better explanation. 
One possible experiment 
is to use a phylogeny of the new diversification model,
simulate its alignment and see how well our
current inference models are at retrieving the phylogeny.
If this difference is small enough, current models apparently
suffice.

\verb;pirouette; is an R package that performs such an experiment,
and is built on \verb;babette; [Bilderbeek \& Etienne, 2018], 
which calls the popular Bayesian inference tool 
BEAST2 [\cite{bouckaert2014beast}]. With \verb;pirouette;, one
can easily measure the error made by Bayesian inference in recovering
any given phylogeny, helping us to valorize new diversification models.

@@

The pipeline is summarized by the following steps, which will be described in detail below:
\begin{enumerate}
    \item from a given phylogeny an alignment is simulated;
    \item from the alignment an inference model is chosen;
    \item the alignment is used as BEAST2 input to infer a posterior distribution of phylogenies;
    \item posterior phylogenies are compared with the given phylogenies to estimate the error we make. The bigger the match, the lower the error;
\end{enumerate}
There is also the option to generate a 'twin tree',
that goes through the same pipeline.

@@

The second step selects an inference model, using the alignment.
The user can specify whether to use the generative model or select a best model
from a set of inference models. 
When selecting the generative model,
the site and clock model used in the alignment simulation are used
in the inference. Because the given phylogeny (on which the alignment is based)
may have followed any tree prior (i.e. speciation model), the user needs
to specify which tree prior is used in the inference. 
When selecting for the best
model, the alignment is used to find the inference model that has the
highest evidence (i.e. marginal likelihood) from a set of inference models.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;babette;, that call the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac; 
inference model selection is not available on Windows.

@@

The fourth step measures the inference error, using the phylogenies in the
Bayesian posterior. These phylogenies are compared to the given
phylogeny using an error statistic, which is the nLTT 
statistic (\cite{janzen2015approximate}) by default,
but also a custom error statistic is supplied, based
on the absolute difference in the gamma statistic [Pybus \& Harvy]. 
Additionally, the user can specify the
proportion of posterior phylogenies to 
discard (i.e. the burn-in), throwing away the first $10\%$
of all phylogenies by default. This burn-in is used to discard
the part of the MCMC chain that has not yet converged to a
representative part of state space.

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
install_beast2()
\end{lstlisting}
For the option to select inference models,
\verb;pirouette; uses the "NS" BEAST2 package [Maturana et al.].
It can be installed from within R, using:

@@

The second step is to select how an inference model is picked.
We just select the generative model, which consists in a collection of 
three elements: a site model, clock model and tree prior.
If we know how the generative model of the alignment, as we did above,
we can re-use that information. We do still need to specify 
a generative tree prior, which is the tree prior underlying the phylogeny.
In this case, we'll pick a (constant-rate) birth-death model:

@@

\richel{Update figure 2 to use idealized tree}

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_2.png}
  \caption{
    The error BEAST2 makes from a phylogeny using the same 
    diversification model as it was generated by
  }
\end{figure}

@@

The second step is to select how an inference model is picked.
We will let the inference model that has the highest evidence be used
in the Bayesian inference. By default, \verb;pirouette; will try
all combinations of site models, clock models and tree priors.

@@

We use all the same settings as the previous model:
phylogeny, alignment parameters, inference model selection,
shared inference parameters and error measuring parameters 
\richel{TODO: add ref to code}

This time, we are interested in creating a twin tree. A twin tree
has the same topology as the given tree, yet its branch lengths follow
an idealized birth-death distribution. Creating a twinning parameter is easy,
as it has sensible default settings:

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_4.png}
  \caption{
    The error BEAST2 makes from a phylogeny 
    picking the best inference model, compared to the background noise?
    Here, the twin column shows the error BEAST2 makes on an idealized
    tree, to measure the noise, which is the minimal error.
  }
\end{figure}

>>>>

The development of new powerful inference tools, 
such as BEAST [\cite{drummond2007beast}] or MrBayes [\cite{huelsenbeck2001mrbayes}], 
provides us the possibility to build phylogenetic trees 
from genetic material extracted from extant organisms.
This has constituted an important step forward 
in our understanding on how species 
evolve.
Such tools have been increasingly exploited to hypothesize and test what are the main drivers
and modes for diversification processes.

There are plenty of diversification models: 
constant-birth [Yule, 19..],
birth-death [\cite{nee1994reconstructed}], 
time-dependent [Rabosky D., LASER package?], 
diversity dependent [\cite{etienne2011diversity}], 
protracted birth-death [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}],
multiple-birth death [Laudanno, Bilderbeek, Haegeman \& Etienne, in preparation],
a rate shifting birth-death model [the SLS model. Laudanno, Haegeman \& Etienne, in preparation]
punctuated equilibrium (as known as "early burst") [?Harmon \giovanni{is it this one? https://onlinelibrary.wiley.com/doi/full/10.1111/j.1558-5646.2010.01025.x}],
dependent on a binary trait [\cite{maddison2007estimating}], 
multiple state trait [\cite{fitzjohn2012diversitree}],
concealed state [\cite{beaulieu2016detecting}] 
or even a combination of multiple concealed and observable states [\cite{herrera2018detecting}].

Such models usually rely on the assumption 
that a particular diversification process 
can be explained mainly by a specific mechanism.
Such a mechanism is built into a likelihood function, 
which provides the likelihood of the parameters, given the data.
By maximizing it, it is possible 
to estimate the what are the best values for model parameters. 
It is a standard practice in assessing such models,
to verify the level of accuracy they can achieve
on simulated phylogenies, of which full knowledge is given.

Diversification models are getting more complex over time. There are many
reasons for this. One reason is the increase in computing power,
which allows this heightened complexity possible.
Another reason is to explore a process that are not yet considered by currently existing diversification models. An example of this
is the multiple-birth model [Laudanno et al., in preparation], 
that is the first model to allow simultaneous speciations to co-occur.

However, whenever a new model is proposed there is always a trade-off between the cost of increased complexity versus the benefits that it can provide.
Therefore desirable models have to be able to disentangle the intrinsic complexity of a system, 
highlight the main reasons of its behaviour and avoid unnecessary complexity. Furthermore developing models can cost time, human and computational resources.

\iffalse
One of the purposes of modelling a complex systems
is to hypothesize the main driving factors of a process 
and trying to prove that they are indeed able 
to explain the main features of the phenomenon you want to explain.
Due to this, there are diminishing returns on adding complexity
to a model: the job of a modeller is to try to
disentangle the intrinsic complexity of a system 
highlighting the main reasons of its behaviour. 
In such a light you don't want your model to be unnecessary complex.
Additionally, developing a likelihood model takes time, 
human resources, fundings and can strain computational resources.
\fi

For this reason we believe it is important to have techniques to quantitatively assess the advancement provided by new models, in comparison to the existing ones.

\iffalse
When assuming that a new diversification model is worth the
(philosophical) effort, there are many ways to measure
if the new model indeed gives a better explanation. 
\fi

To achieve this, one possible experiment
is to simulate a phylogeny according to the mechanisms proposed by the new diversification model, then simulate its alignment and measure how well current inference models perform at retrieving the original phylogeny.
If this difference is small enough, current models apparently
suffice.

\verb;pirouette; is an R package that performs such an experiment,
and is built on \verb;babette; [Bilderbeek \& Etienne, 2018], 
which calls the popular Bayesian inference tool 
BEAST2 [\cite{bouckaert2014beast}]. With \verb;pirouette;, one
can easily measure the error made by Bayesian inference in recovering
any given phylogeny, helping us to validate new diversification models.

@@

The pipeline is summarized by the following steps, which will be described in detail below:
\begin{enumerate}
    \item from a given phylogeny an alignment is simulated;
    \item from the alignment an inference model is chosen;
    \item the alignment is used as BEAST2 input to infer a posterior distribution of phylogenies;
    \item posterior phylogenies are compared with the given original phylogeny to estimate the error we make. The bigger the match, the lower the error;
\end{enumerate}
\giovanni{Here it could be useful to add a figure that explains the pipeline, a bit like the one I have on my desk}
There is also the option to generate a 'twin tree',
that goes through the same pipeline.

@@

The second step selects an inference model, using the alignment.
The user can specify whether to use the generative model or select a best model
from a set of inference models. 
When selecting the generative model,
the site and clock model used in the alignment simulation are used
in the inference. Because the given phylogeny (on which the alignment is based)
may have followed any tree prior (i.e. speciation model), the user needs
to specify which tree prior is used in the inference. 
When selecting for the best
model, the alignment is used to find the inference model that has the
highest evidence (i.e. marginal likelihood) from a set of candidate inference models.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;babette;\giovanni{babette or mcbette?}, that calls the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac; 
inference model selection is not available on Windows.

@@

The fourth step measures the inference error, using the phylogenies in the
Bayesian posterior. These phylogenies are compared to the given
original phylogeny using an error statistic, which is the nLTT 
statistic (\cite{janzen2015approximate}) by default,
but also a custom error statistic is supplied, based
on the absolute difference in the gamma statistic [Pybus \& Harvy]. 
Additionally, the user can specify the
proportion of posterior phylogenies to 
discard (i.e. the burn-in), throwing away the first $10\%$
of all phylogenies by default. This burn-in is used to discard
the part of the MCMC chain that has not yet converged to a
representative part of state space.

@@

\begin{lstlisting}[language=R, floatplacement=H, frame=single]
install_beast2()
\end{lstlisting}
For the option to select inference models,
\verb;pirouette; uses the "NS" BEAST2 package [\cite{maturana2018model}].
It can be installed from within R, using:

@@

The second step is to select how an inference model is picked.
We just select the generative model, which consists in a collection of 
three elements: a site model, clock model and tree prior. \giovanni{Here you present the "alignment generative model" and the "inference model". These are general concepts, as they are used in all the experiments. I propose to define them from the start, for example in the "description" section; we can even refer to them in the figure I propose to add, so to make it clearer to the reader. It can be done in this way:
A "model" is always defined as the collection of three elements: a site model, a clock model and a tree prior. We have two kind of models: generative and inference. The first is used to simulate alignments, the second is used to produce a posterior. Generative models always use a (constant-rate) birth-death tree prior.}
If we know how the generative model of the alignment, as we did above,
we can re-use that information. We do still need to specify 
a generative tree prior, which is the tree prior underlying the phylogeny.
In this case, we'll pick a (constant-rate) birth-death model:

@@

\richel{Update figure 2 to use idealized tree}
\giovanni{Check "pirouettepackage" repo. I uploaded a function to create the idealized tree that you might like.}

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_2.png}
  \caption{
    The error BEAST2 makes from a phylogeny using the same 
    diversification model as it was generated by
    \giovanni{If we define generative and inference model from the start we can rephrase this caption in a better way, referring to terms that the reader is already familiar to}
  }
\end{figure}

@@

The second step is to select how an inference model is picked.
We will let the inference model that has the highest evidence to be used
in the Bayesian inference. By default, \verb;pirouette; will try
all combinations of site models, clock models and tree priors.\giovanni{again, if we give a name to this triplet, it could be good to call it by its name. We can still recall its composition in brackets, if we think it could be useful}

@@

The following settings are the same as in the previous section:
phylogeny, alignment parameters, inference model selection,
shared inference parameters and error measuring parameters 
\richel{TODO: add ref to code}

This time, we are interested in creating a twin tree. A twin tree
has the same topology as the given tree, yet its branch lengths follow
an idealized \giovanni{I don't think this is true. A twin tree is generated randomly. If we use the method "maxlikelihood" (see the function "createbdtree") the tree will be chosen in such a way it will maximizing the likelihood with the parameters we estimated from the original tree, but I would not call it "idealized". Furthermore I think it's not obvious to use the "maxlikelihood" method, as in general the "ideal" scenario should be established using a distribution of trees. A single tree is always the result of a stochastic process.} birth-death distribution. Creating a twinning parameter is easy,
as it has sensible default settings:

@@

\begin{figure}[h]
  \includegraphics[width=\textwidth]{figure_4.png}
  \caption{
    The error BEAST2 makes from a phylogeny 
    picking the best inference model, compared to the background noise?
    Here, the twin column shows the error BEAST2 makes on an idealized
    tree, to measure the noise, which is the minimal error. \giovanni{again, I don't believe that using an "idealized" tree is the best thing to do. I would rather use more "true trees".}
  }
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
\giovanni{We need to discuss the results we got from the 3 experiments.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
